---
title: "R Notebook"
output: html_notebook
---

```{r}
# code to load packages
library(tidyverse)
library(tidymodels)
library(knitr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(corrplot)

#baseR
```

```{r}
auto <- read_csv("data/datagenu.csv")

#Set hms as numeric 
auto$hm1 <- as.numeric(as.character(auto$hm1))
auto$hm2 <- as.numeric(as.character(auto$hm2))
auto$hm3 <- as.numeric(as.character(auto$hm3))
auto$hm4 <- as.numeric(as.character(auto$hm4))
auto$hm5 <- as.numeric(as.character(auto$hm5))

#Set hpps as numeric 
auto$hpp1 <- as.numeric(as.character(auto$hpp1))
auto$hpp2 <- as.numeric(as.character(auto$hpp2))
auto$hpp3 <- as.numeric(as.character(auto$hpp3))
auto$hpp4 <- as.numeric(as.character(auto$hpp4))
auto$hpp5 <- as.numeric(as.character(auto$hpp5))
```


```{r}
#remove NAs
autoclean <- auto %>%
  filter(complete.cases(hm1, hm2, hm3, hm4, hm5, hpp1, hpp2, hpp3, hpp4, hpp5))

```

```{r}
#avg hand metatarsals for each higher taxa

hm_avg <- autoclean |>
  group_by(aagroup) |>
  summarize(average_hm1 = mean(hm1, na.rm = TRUE),
            average_hm2 = mean(hm2, na.rm = TRUE),
            average_hm3 = mean(hm3, na.rm = TRUE),
            average_hm4 = mean(hm4, na.rm = TRUE),
            average_hm5 = mean(hm5, na.rm = TRUE)
            )
print(hm_avg)
```
```{r}
#avg proximal phalanges for each higher taxa

hpp_avg <- autoclean |>
  group_by(aagroup) |>
  summarize(average_hpp1 = mean(hpp1, na.rm = TRUE),
            average_hpp2 = mean(hpp2, na.rm = TRUE),
            average_hpp3 = mean(hpp3, na.rm = TRUE),
            average_hpp4 = mean(hpp4, na.rm = TRUE),
            average_hpp5 = mean(hpp5, na.rm = TRUE)
            )
print(hpp_avg)
```

```{r}
#get total digit lengths 

autoclean$total_digit1 <- autoclean$hm1 + autoclean$hpp1
autoclean$total_digit2 <- autoclean$hm2 + autoclean$hpp2
autoclean$total_digit3 <- autoclean$hm3 + autoclean$hpp3
autoclean$total_digit4 <- autoclean$hm4 + autoclean$hpp4
autoclean$total_digit5 <- autoclean$hm5 + autoclean$hpp5

head(autoclean)

```
```{r}
#AVG DIGIT LENGTHS

avg_digit_lengths <- hm_avg |>
  full_join(hpp_avg, by = "aagroup") |>

  mutate(
      avg_digit1 = hm_avg$average_hm1 + hpp_avg$average_hpp1,
    avg_digit2 = hm_avg$average_hm2 + hpp_avg$average_hpp2,
    avg_digit3 = hm_avg$average_hm3 + hpp_avg$average_hpp3,
    avg_digit4 = hm_avg$average_hm4 + hpp_avg$average_hpp4,
    avg_digit5 = hm_avg$average_hm5 + hpp_avg$average_hpp5
    ) |>
    select(aagroup, avg_digit1, avg_digit2, avg_digit3, avg_digit4, avg_digit5)

avg_digit_lengths
```

```{r}
#EXPLORING COVARIATION
# BIPLOTS FOR ALL DIGITS IN DATASET 

#to calc rsquared 
# r_squared <- function(x, y) {
#   model <- lm(y ~ x)
#   summary(model)$r.squared
# }
# 
# pairs <- list(
#   c("total_digit1", "total_digit2"),
#   c("total_digit1", "total_digit3"),
#   c("total_digit1", "total_digit4"),
#   c("total_digit1", "total_digit5"),
#   c("total_digit2", "total_digit3"),
#   c("total_digit2", "total_digit4"),
#   c("total_digit2", "total_digit5"),
#   c("total_digit3", "total_digit4"),
#   c("total_digit3", "total_digit5"),
#   c("total_digit4", "total_digit5")
# )
# 
# for (pair in pairs) {
#   x <- autoclean[[pair[1]]]
#   y <- autoclean[[pair[2]]]
#   r_squared <- calculate_r_squared(x, y)
#   p <- ggplot(autoclean, aes_string(x = pair[1], y = pair[2])) +
#     geom_point() +
#     labs(x = pair[1], y = pair[2], title = paste("Biplot of", pair[1], "and", pair[2]), subtitle = paste("R^2 =", r_squared)) +
#     theme_minimal()
#   print(p)
# }

```
```{r}
# #CORRELATION MATRIX 
# correlation_matrix <- cor(autoclean[, c("total_digit1", "total_digit2", "total_digit3", "total_digit4", "total_digit5")])
# print(correlation_matrix)
# 
# corrplot(correlation_matrix, method = "color", addCoef.col = "black", tl.col = "black", tl.srt = 45)
```
```{r}

# Function to calculate R^2
# calc_r_squared <- function(x, y) {
#   model <- lm(y ~ x)
#   summary(model)$r.squared
# }
# 
# # Create pairs of digit lengths
# pairs <- list(
#   c("total_digit1", "total_digit2"),
#   c("total_digit1", "total_digit3"),
#   c("total_digit1", "total_digit4"),
#   c("total_digit1", "total_digit5"),
#   c("total_digit2", "total_digit3"),
#   c("total_digit2", "total_digit4"),
#   c("total_digit2", "total_digit5"),
#   c("total_digit3", "total_digit4"),
#   c("total_digit3", "total_digit5"),
#   c("total_digit4", "total_digit5")
# )
# 
# # Get unique taxa
# unique_taxa <- unique(autoclean$aagroup)
# 
# # for each higher taxa 
# for (taxon in unique_taxa) {
#   # filter for the current higher taxon
#   taxa_data <- autoclean[autoclean$aagroup == taxon, ]
#   
#   # going thru each digit pair 
#   for (pair in pairs) {
#     x <- taxa_data[[pair[1]]]
#     y <- taxa_data[[pair[2]]]
#     
#     # get r^2
#     r_squared <- calc_r_squared(x, y)
#     
#     # actual plot
#     p <- ggplot(taxa_data, aes_string(x = pair[1], y = pair[2])) +
#       geom_point(color = "#0096FF", alpha = 0.6, size = 3) +
#       labs(
#         x = pair[1],
#         y = pair[2],
#         title = paste("Biplot of", pair[1], "and", pair[2], "for", taxon),
#         subtitle = paste("R^2 =", round(r_squared, 3))
#       ) +
#       theme_minimal() +
#       facet_wrap(~ `Sex/age`) +
#       theme(
#         plot.title = element_text(size = 12, face = "bold", color = "#0096FF"),
#         plot.subtitle = element_text(size = 10, color = "black"),
#         axis.title.x = element_text(size = 10, face = "bold"),
#         axis.title.y = element_text(size = 10, face = "bold"),
#         panel.background = element_rect(fill = "#FFF1F3"),  
#         panel.grid.major = element_line(color = "white"), 
#         panel.grid.minor = element_line(color = "white")
#       )
#     print(p)
#   }
# }

```

```{r}

# Function to calculate R^2
#calc_r_squared <- function(x, y) {
  #model <- lm(y ~ x)
 # summary(model)$r.squared
#}

# Create pairs of digit lengths
#pairs <- list(
  #c("total_digit1", "total_digit2"),
  #c("total_digit1", "total_digit3"),
  #c("total_digit1", "total_digit4"),
  #c("total_digit1", "total_digit5"),
  #c("total_digit2", "total_digit3"),
  #c("total_digit2", "total_digit4"),
  #c("total_digit2", "total_digit5"),
  #c("total_digit3", "total_digit4"),
  #c("total_digit3", "total_digit5"),
 # c("total_digit4", "total_digit5")
#)

# Get unique taxa
#unique_taxa <- unique(autoclean$aagroup)

# for each higher taxa 
#for (taxon in unique_taxa) {
  # filter for the current higher taxon
  #taxa_data <- autoclean[autoclean$aagroup == taxon, ]
 
  # going thru each digit pair 
  #for (pair in pairs) {
    #x <- taxa_data[[pair[1]]]
    #y <- taxa_data[[pair[2]]]
    
    # get r^2
   # r_squared <- calc_r_squared(x, y)
    
    # actual plot
  #  p <- ggplot(taxa_data, aes_string(x = pair[1], y = pair[2])) +
      #geom_point(color = "#0096FF", alpha = 0.6, size = 3) +
      #labs(
        #x = pair[1],
        #y = pair[2],
        #title = paste("Biplot of", pair[1], "and", pair[2], "for", taxon),
       # subtitle = paste("R^2 =", round(r_squared, 3))
      #) +
      #theme_minimal() +
      #theme(
        #plot.title = element_text(size = 12, face = "bold", color = "#0096FF"),
        #plot.subtitle = element_text(size = 10, color = "black"),
        #axis.title.x = element_text(size = 10, face = "bold"),
       # axis.title.y = element_text(size = 10, face = "bold"),
      #  panel.background = element_rect(fill = "#FFF1F3"),  
     #   panel.grid.major = element_line(color = "white"), 
    #    panel.grid.minor = element_line(color = "white")
   #   )
  #  print(p)
 # }

#}
 
```

```{r}
#CONFUSION MATRICES FOR EACH AAGROUP

# Get unique taxa
#unique_taxa <- unique(autoclean$Taxon)

# for each higher taxa 
#for (taxon in unique_taxa) {
  # filter for the current higher taxon
 # taxa_data <- autoclean[autoclean$Taxon == taxon, ]
  
  #empty matrix 
  # r_squared_matrix <- matrix(NA, nrow = 5, ncol = 5)
   # rownames(r_squared_matrix) <- colnames(r_squared_matrix) <- c("total_digit1", "total_digit2", "total_digit3", "total_digit4", "total_digit5")
    
 #for (pair in pairs) {
  #  x <- taxa_data[[pair[1]]]
   # y <- taxa_data[[pair[2]]]
    
    # get r^2
    #r_squared <- calc_r_squared(x, y)
    
    # Store R^2 in the matrix
    #r_squared_matrix[pair[1], pair[2]] <- r_squared
    #r_squared_matrix[pair[2], pair[1]] <- r_squared
 #}
  #    print(paste("R^2 matrix for taxon:", taxon))
   # print(r_squared_matrix)
    
    # Plot R^2 matrix as a heatmap
    #corrplot(r_squared_matrix, method = "color", addCoef.col = "black", tl.col = 
     #          "black", tl.srt = 45, title = paste("R^2 Matrix for", taxon), mar=c(0,0,1,0))
  
#}
  
```
1. For lorisiform:
- 5&2, 1&2, 5&3, 4&2 are least correlated. 
-  1&4, 3&2, 5&4, 3&4, 5&1 are most correlated 

2. For lemurid: 
- Everything is pretty correlated --> probably because lemurid represents a large number of the data in its groupings
- 3&1, 3&2, 3&5 least correlated 
- 4&3, 4&5 have highest correlation
- future plan: potentionally split by taxon next

3. For OWM: 
- 1&3, 1&4, 1&2, 1&5 are least correlated
- everything else is pretty correlated (higher than .94)
- Odd pattern? Why are only digits 1 not correlated with anything?

Overall:
- 4&5, 3&4, 1&5 have highest correlation 
- 3&5 least correlated 

```{r}
#unique_taxa <- unique(autoclean$aagroup)

#for (taxon in unique_taxa) {
 # taxa_data <- autoclean[autoclean$aagroup == taxon, ]
  #if (nrow(taxa_data) > 2 && all(complete.cases(taxa_data[, c("total_digit1", 
   #                                                           "total_digit2", 
    #                                                          "total_digit3", 
     #                                                         "total_digit4", 
      #                                                        "total_digit5")]))) {
  #correlation_matrix <- cor(taxa_data[, c("total_digit1", "total_digit2", 
   #                                       "total_digit3", "total_digit4", 
    #                                      "total_digit5")])
  #print(taxon)
  #print(correlation_matrix)
  #corrplot(correlation_matrix, method = "color", addCoef.col = "black", tl.col = 
   #          "black", tl.srt = 45,   title=taxon,    mar=c(0,0,1,0) )
  #}
#}
```


```{r}
# autoclean_grouped <- autoclean %>%
#   group_by(genus) %>%  # Group by Taxon
#   summarize(
#     mc_total = sum(hm1, hm2, hm3, hm4, hm5, na.rm = TRUE), # Sum of metacarpals within group
#     ph_total = sum(hpp1, hpp2, hpp3, hpp4, hpp5, na.rm = TRUE), # Sum of phalanges within group
#     mc1_sum = sum(hm1, na.rm = TRUE),
#     mc2_sum = sum(hm2, na.rm = TRUE),
#     mc3_sum = sum(hm3, na.rm = TRUE),
#     mc4_sum = sum(hm4, na.rm = TRUE),
#     mc5_sum = sum(hm5, na.rm = TRUE),
#     ph1_sum = sum(hpp1, na.rm = TRUE),
#     ph2_sum = sum(hpp2, na.rm = TRUE),
#     ph3_sum = sum(hpp3, na.rm = TRUE),
#     ph4_sum = sum(hpp4, na.rm = TRUE),
#     ph5_sum = sum(hpp5, na.rm = TRUE),
#     .groups = "drop" # Prevent warning about grouped data
#   ) %>%
#   mutate(
#     mc_total = ifelse(mc_total == 0, NA, mc_total),  # Handle groups with 0 sum
#     ph_total = ifelse(ph_total == 0, NA, ph_total)   # Handle groups with 0 sum
#   )
# 
# # Reattach aagroup
# aagroup_mapping <- autoclean %>%
#   select(genus, aagroup) %>%
#   distinct()  # Remove duplicates
# 
# autoclean_grouped <- autoclean_grouped %>%
#   left_join(aagroup_mapping, by = "genus")
# 
# head(autoclean_grouped)
# 
# # calculate the geomeans for each metacarpal and phalange (group-level)
# autoclean_grouped <- autoclean_grouped %>%
#   mutate(
#     mc1_geomean = mc1_sum / mc_total,
#     mc2_geomean = mc2_sum / mc_total,
#     mc3_geomean = mc3_sum / mc_total,
#     mc4_geomean = mc4_sum / mc_total,
#     mc5_geomean = mc5_sum / mc_total,
#     ph1_geomean = ph1_sum / ph_total,
#     ph2_geomean = ph2_sum / ph_total,
#     ph3_geomean = ph3_sum / ph_total,
#     ph4_geomean = ph4_sum / ph_total,
#     ph5_geomean = ph5_sum / ph_total
#   )
# 
# # the relevant columns
# autoclean_new <- autoclean_grouped %>%
#   select(
#     genus,
#     aagroup,
#     mc1_geomean, mc2_geomean, mc3_geomean, mc4_geomean, mc5_geomean,
#     ph1_geomean, ph2_geomean, ph3_geomean, ph4_geomean, ph5_geomean
#   )
# 
# # save dataset
# write.csv(autoclean_new, "fin_aageomean.csv", row.names = FALSE)
# 
# 
# getwd()  # Print the working directory where the file is saved
# print(head(autoclean_new))  # Print the first few rows of the new data
# 

```


```{r}
install.packages("ggplot")
install.packages("ggplot2")
install.packages("ggrepel")  # for better label positioning
install.packages("magrittr") # package installations are only needed the first time you use it
install.packages("dplyr")    # alternative installation of the %>%
```


```{r}
library(ggplot2)
library(ggrepel)
library(magrittr) # needs to be run every time you start R and want to use %>%
library(dplyr)    # alternatively, this also loads %>%
library(tidyr)
```



```{r}
my_data <- read.csv("data/geocleandd.csv", header = TRUE)

# Replace #DIV/0! with NA
my_data[my_data == "#DIV/0!"] <- NA

# Convert all columns except for taxonomic info to numeric
my_data[, 4:ncol(my_data)] <- lapply(my_data[, 4:ncol(my_data)], as.numeric)

# Check for missing values
summary(my_data)

```

```{r}

# Load the cleaned dataset

# Load necessary libraries
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggrepel")) install.packages("ggrepel")
if (!require("dplyr")) install.packages("dplyr")
if (!require("tidyr")) install.packages("tidyr")

library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
```

```{r}

##############################
# Run PCA
##############################

# Ensure genus and aagroup columns are intact
genus_column <- my_data$genus
aagroup_column <- my_data$aagroup

# Select numeric geomean columns for PCA
numeric_data <- my_data %>%
  select(mc1_geo, mc2_geo, mc3_geo, mc4_geo, mc5_geo,
         mPP1.mIP1_geo, mPP2.mIP2_geo, mPP3.mIP3_geo, mPP4.mIP4_geo, mPP5.mIP5_geo)

# Retain row indices and remove rows with NA values
numeric_data_clean <- numeric_data %>% na.omit()
clean_indices <- rownames(numeric_data_clean)

# Perform PCA
pca_result <- prcomp(numeric_data_clean, center = TRUE, scale. = TRUE)

# Align taxonomic information with PCA results
pca_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  genus = genus_column[rownames(my_data) %in% clean_indices],
  aagroup = aagroup_column[rownames(my_data) %in% clean_indices]
)

# Check PCA data
head(pca_data)
print(unique(pca_data$genus))  # Should show genera present in the cleaned dataset

# View PCA loadings
pca_loadings <- pca_result$rotation

##############################
# Visualize PCA Results
##############################

# Create hulls for taxonomic groups
hull_data <- pca_data %>%
  group_by(genus) %>%
  slice(chull(PC1, PC2))  # Convex hull for each group

# Create a biplot
library(ggplot2)
library(ggrepel)

biplot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = genus)) +
  geom_polygon(data = hull_data, alpha = 0.3, aes(group = genus), show.legend = FALSE) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(aes(label = genus), size = 3, max.overlaps = Inf) +
  labs(title = "PCA Biplot with Geomeans",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()

# Print the plot
print(biplot)

##############################
# Analyze PCA Loadings
##############################

# Print loadings sorted by variance contribution
loadings_variation <- as.data.frame(pca_loadings) %>%
  mutate(Variance = rowSums(pca_loadings^2)) %>%
  arrange(desc(Variance))

print("Loadings Sorted by Variance Contribution:")
print(loadings_variation)

# Visualize PCA loadings
plot_pca_loadings <- function(loadings_df, pc_number) {
  ggplot(loadings_df, aes_string(x = "rownames(loadings_df)", y = paste0("PC", pc_number))) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(title = paste("PCA Loadings for PC", pc_number),
         x = "Variables",
         y = "Loadings") +
    theme_minimal()
}

# Plot loadings for PC1 and PC2
plot_pca_loadings(loadings_variation, 1)
plot_pca_loadings(loadings_variation, 2)

##############################
# Variance Explained by Each PC
##############################

# Calculate percent variance explained by each principal component
percent_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100

# Create a variance data frame
variance_df <- data.frame(
  Principal_Component = paste0("PC", 1:length(percent_variance)),
  Percent_Variance = round(percent_variance, 2)
)

print("Percent Variance Explained:")
print(variance_df)


```




```{r}
##############################
# Run PCA
##############################

# Ensure genus and aagroup columns are intact
genus_column <- my_data$genus
aagroup_column <- my_data$aagroup

# Select numeric geomean columns for PCA
numeric_data <- my_data %>%
  select(mc1_geo, mc2_geo, mc3_geo, mc4_geo, mc5_geo,
         mPP1.mIP1_geo, mPP2.mIP2_geo, mPP3.mIP3_geo, mPP4.mIP4_geo, mPP5.mIP5_geo)

# Retain row indices and remove rows with NA values
numeric_data_clean <- numeric_data %>% na.omit()
clean_indices <- rownames(numeric_data_clean)

# Perform PCA
pca_result <- prcomp(numeric_data_clean, center = TRUE, scale. = TRUE)

# Align taxonomic information with PCA results
pca_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  aagroup = aagroup_column[rownames(my_data) %in% clean_indices]  # Align using cleaned indices
)

# Check PCA data
head(pca_data)
print(unique(pca_data$aagroup))  # Should show the aagroups present in the cleaned dataset

# View PCA loadings
pca_loadings <- pca_result$rotation

##############################
# Visualize PCA Results
##############################

# Create hulls for taxonomic groups (grouped by aagroup)
hull_data <- pca_data %>%
  group_by(aagroup) %>%
  slice(chull(PC1, PC2))  # Convex hull for each group

# Create a biplot grouped by aagroup
library(ggplot2)
library(ggrepel)

biplot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = aagroup)) +
  geom_polygon(data = hull_data, alpha = 0.3, aes(group = aagroup), show.legend = FALSE) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(aes(label = aagroup), size = 3, max.overlaps = Inf) +
  labs(title = "PCA Biplot Grouped by Aagroup",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()

# Print the plot
print(biplot)

##############################
# Analyze PCA Loadings
##############################

# Print loadings sorted by variance contribution
loadings_variation <- as.data.frame(pca_loadings) %>%
  mutate(Variance = rowSums(pca_loadings^2)) %>%
  arrange(desc(Variance))

print("Loadings Sorted by Variance Contribution:")
print(loadings_variation)

# Visualize PCA loadings
plot_pca_loadings <- function(loadings_df, pc_number) {
  ggplot(loadings_df, aes_string(x = "rownames(loadings_df)", y = paste0("PC", pc_number))) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(title = paste("PCA Loadings for PC", pc_number),
         x = "Variables",
         y = "Loadings") +
    theme_minimal()
}

# Plot loadings for PC1 and PC2
plot_pca_loadings(loadings_variation, 1)
plot_pca_loadings(loadings_variation, 2)

##############################
# Variance Explained by Each PC
##############################

# Calculate percent variance explained by each principal component
percent_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100

# Create a variance data frame
variance_df <- data.frame(
  Principal_Component = paste0("PC", 1:length(percent_variance)),
  Percent_Variance = round(percent_variance, 2)
)

print("Percent Variance Explained:")
print(variance_df)

```



